Supporting Information for "Can we distinguish model of selective interactions using linkage disequilibrium?

### 2021-03-08 Manuscript.Rmd
edits

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting patterns of allele frequencies and correlations between pairs
of mutations. A number of recent studies have used patterns of linkage
disequilibrium between selected variants to test for selective interference and
epistatic interactions, with some disagreement over interpretation of
observations. Interpreting this data is hindered by the relative lack of
analytic or even numerical expectations for patterns of variation between pairs
of loci under the combined effects of selection, dominance, epistasis, and
demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection, with arbitrary
epistasis and dominance, and variable recombination and population sizes. I use
this to explore how different models of epistasis and dominance affect expected
signed LD, including for non-steady-state demography relevant to human
populations. Finally, I use whole-genome sequencing data from humans to asses
how well we can differentiate modes of selective interactions in practice. I
find that observations of positive LD between missense mutations within genes
is driven by strong positive correlations between pairs of mutations that fall
within the same conserved domain, pointing to compensatory mutations or
antagonistic epistasis as the prevailing mode of interaction within but not
outside of conserved genic elements. The heterogeneous landscape of both
strength of selection and selective interactions within protein-coding genes
calls for more refined inferences of the joint distribution of fitness and
interactive effects, and the methods presented here should prove useful in that
pursuit.

@@

In their foundational paper, @Hill1966-gv recognized that linked selected
variants reciprocally impede the efficacy of selection at each locus. Linked
selection reduces the fixation probability of advantageous mutations and
increases that of deleterious mutations, compared to expectations under
single-locus models [@Birky1988-jm]. Allele frequency dynamics of linked
selected variants are also predicted to deviate from expectations without
interference. Under a multiplicative fitness model, where the fitness reduction
of a haplotype carrying multiple deleterious variants is equal to the product
of the fitness reduction of each mutation independently, we should expect to
see those mutation segregate on different haplotypes more often than together,
leading to negative, or repulsion, linkage disequilibrium (LD), although the
extent of LD depends non-trivially on the strength of selection and probability
of recombination separating loci [@Hill1966-gv; @McVean2000-ox].

Non-additive effects, including dominance (i.e., interactions *within* loci)
and epistasis (interactions *between* loci), further complicate our
evolutionary models. A large fraction of nonsynonymous coding mutations are
thought to be at least partially recessive [@Agrawal2011-wb; @Huber2018-cp],
with average levels of dominance correlating with strength of selection
[@Kacser1981-nc], and dominance plays an important role in shaping expected
equilibrium allele frequencies and mutation load of strongly damaging disease
mutations [@Clark1998-kq]. On the other hand, epistasis differentially impacts
the deleterious load for asexually and sexually reproducing organisms
[@Kimura1966-gc; @Kondrashov1995-va], has been invoked as an explanation for
the evolutionary advantage of sex [@Kondrashov1982-sf; @Charlesworth1990-kw;
@Barton1998-lu], and can drive incompatibilities that lead to postzygotic
isolation during the process of speciation [@Turelli2000-kz]. Within
populations, epistasis is known to cause signed LD to deviate dramatically from
zero [@Charlesworth1990-kw; @Kondrashov1995-va]. However, despite dominance
being appreciated as important factors shaping linked variation
[@Turelli2000-kz; @Zhao2016-bb] and the evolutionary importance of epistatic
interactions, we currently lack models for predicting correlation patterns
between linked mutations under general selection models.

@@

In most natural populations, directed mutational studies are not possible, and
we must turn to population genetic approaches to infer selective interactions
between observed mutations. Motivated by theory that linked negatively selected
mutations will display negative LD due to interference [@Hill1966-gv], and that
epistasis will drive expected LD away from zero, a number of recent studies
have used patterns of LD within classes of putatively selected variants to
infer modes of selective interactions. In a notable study from
@Callahan2011-ac, pairs of nonsynonymous mutations were found to cluster more
than expected along lineages in the *Drosophilid* species complex, and that
those clustered mutations tended to preserve the charge of the protein and were
in positive LD compared to pairs involving synonymous mutations, suggesting
that compensatory nonsynonymous variants are tolerated and maintained. More
recently, @Taverner2020-lk replicated this finding across a diverse set of
genera, showing that such epistatic interactions are important for protein
evolution.

@@

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function variants,
their rarity and low frequency means that statistical measurement of LD and
other diversity measures are very noisy. Second, comparisons are based on
theory that are confined to limiting and simplistic assumptions, including
steady-state demography, simple selection and interaction models, or unlinked
loci. To generate predictions under more complex models, we rely on expensive
forward simulations. Such simulations can help build intuition and test
inference methods, but do not efficiently provide expectations for quantities
of interest across the range of relevant parameters. Analytical and numerical
methods for expected haplotype frequencies and LD under general selective
interaction models are thus crucial for interpreting patterns of variation
observed in data.

@@

I first consider the case of additive selection and epistasis, followed by
dominance acting within loci but no epistatic effects, and then describe a
general diploid selection model and consider simple gene-based dominance
effects. In what follows I focus on parameters where the strengths of selection
and dominance at each locus are equal, but note that the methods and
implementation allow for arbitrary and unequal selection and dominance at the
two loci.

@@

In a single randomly mating population, neutral mutations are expected to have
average signed LD of zero, but across all populations analyzed here, we observe
positive LD between synonymous mutations. While selection on some subset of
synonymous variants is possible, it should be much weaker on average than
between missense mutations, and any interference between selected synonymous
variants should lead to negative LD. @Sandler2020-vk proposed that spatial
population structure is responsible for the increase of LD for both synonymous
and missense variants, arguing that every natural population lives in some
geography that will necessarily induce some structure, and they used forward
simulations to demonstrate this effect. However, this increase in LD, at least
in simulations of spatial structure and migration, typically extends over larger
distances separating base-pairs than we observe in the data. Instead, positive
LD between synonymous mutations decays to zero within just a few hundred
base-pairs (Figures \@ref(fig:lddecay) and \@ref(fig:lddecaydomain)).

>>>>

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting patterns of allele frequencies and correlations between pairs
of mutations. A number of recent studies have used patterns of linkage
disequilibrium between selected variants to test for selective interference and
epistatic interactions, with some disagreement over interpretation of
observations. Interpreting this data is hindered by the relative lack of
analytic or even numerical expectations for patterns of variation between pairs
of loci under the combined effects of selection, dominance, epistasis, and
demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection, with arbitrary
epistasis and dominance, and variable recombination and population sizes. I use
this to explore how different models of epistasis and dominance affect expected
signed LD, including for non-steady-state demography relevant to human
populations. Finally, I use whole-genome sequencing data from humans to asses
how well we can differentiate modes of selective interactions in practice. I
find that positive LD between missense mutations within genes is driven by
strong positive correlations between pairs of mutations that fall within the
same conserved domain, pointing to compensatory mutations or antagonistic
epistasis as the prevailing mode of interaction within but not outside of
conserved genic elements. The heterogeneous landscape of both strength of
selection and selective interactions within protein-coding genes calls for more
refined inferences of the joint distribution of fitness and interactive
effects, and the methods presented here should prove useful in that pursuit.

@@

In their foundational paper, @Hill1966-gv recognized that linked selected
variants reciprocally impede the efficacy of selection at each locus. Linked
selection reduces the fixation probability of advantageous mutations and
increases that of deleterious mutations compared to expectations under
single-locus models [@Birky1988-jm]. Allele frequency dynamics of linked
selected variants are also predicted to deviate from expectations without
interference. Under a multiplicative fitness model, where the fitness reduction
of a haplotype carrying multiple deleterious variants is equal to the product
of the fitness reduction of each mutation independently, we should expect to
see those mutation segregate on different haplotypes more often than together,
leading to negative, or repulsion, linkage disequilibrium (LD), although the
extent of LD depends non-trivially on the strength of selection and probability
of recombination separating loci [@Hill1966-gv; @McVean2000-ox].

Non-additive effects, including dominance (i.e., interactions *within* loci)
and epistasis (interactions *between* loci), further complicate our
evolutionary models. A large fraction of nonsynonymous coding mutations are
thought to be at least partially recessive [@Agrawal2011-wb; @Huber2018-cp],
with average levels of dominance correlating with strength of selection
[@Kacser1981-nc], and dominance plays an important role in shaping expected
equilibrium allele frequencies and the mutation load of strongly damaging
disease mutations [@Clark1998-kq]. On the other hand, epistasis differentially
impacts the deleterious load for asexually and sexually reproducing organisms
[@Kimura1966-gc; @Kondrashov1995-va], has been invoked as an explanation for
the evolutionary advantage of sex [@Kondrashov1982-sf; @Charlesworth1990-kw;
@Barton1998-lu], and can drive incompatibilities that lead to postzygotic
isolation during the process of speciation [@Turelli2000-kz]. Within
populations, epistasis is known to cause signed LD to deviate dramatically from
zero [@Charlesworth1990-kw; @Kondrashov1995-va]. However, despite appreciation
of the effect of dominance on linked variation [@Turelli2000-kz; @Zhao2016-bb]
and the evolutionary importance of epistatic interactions, we currently lack
models for predicting correlation patterns between linked mutations under
general selection models.

@@

Directed mutational studies are not possible in most natural populations, and
we must turn to population genetic approaches to infer selective interactions
between observed mutations. Motivated by theory that linked negatively selected
mutations will display negative LD due to interference [@Hill1966-gv], and that
epistasis will drive expected LD away from zero, a number of recent studies
have used patterns of LD within classes of putatively selected variants to
infer modes of selective interactions. In a notable study from
@Callahan2011-ac, pairs of nonsynonymous mutations were found to cluster more
than expected along lineages in the *Drosophilid* species complex, and that
those clustered mutations tended to preserve the charge of the protein and were
in positive LD compared to pairs involving synonymous mutations, suggesting
that compensatory nonsynonymous variants are tolerated and maintained. More
recently, @Taverner2020-lk replicated this finding across a diverse set of
genera, showing that such epistatic interactions are important for protein
evolution.

@@

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function variants,
their rarity and low frequency means that statistical measurement of LD and
other diversity measures are very noisy. Second, comparisons are based on
theory with limiting assumptions, such as steady-state demography, simple
selection and interaction models, or unlinked loci. To generate predictions
under more complex models, we rely on expensive forward simulations. Such
simulations can help build intuition and test inference methods, but do not
efficiently provide expectations for quantities of interest across the range of
relevant parameters. Analytical and numerical methods for expected haplotype
frequencies and LD under general selective interaction models are thus crucial
for interpreting patterns of variation observed in data.

@@

I first consider the case of additive selection and epistasis, followed by
dominance acting within loci but no epistatic effects, and then describe a
general diploid selection model and consider simple gene-based dominance
effects. In what follows I focus on parameters where the strengths of selection
and dominance at each locus are equal, but note that the methods presented here
allow for arbitrary and unequal selection and dominance at the two loci.

@@

In a single randomly mating population, neutral mutations are expected to have
average signed LD of zero, but across all populations analyzed here, LD between
synonymous mutations is positive. While selection on some subset of
synonymous variants is possible, it should be much weaker on average than
between missense mutations, and any interference between selected synonymous
variants should lead to negative LD. @Sandler2020-vk proposed that spatial
population structure may be responsible for the increase of LD for both synonymous
and missense variants. They argued that every natural population lives in some
geography that will necessarily induce some structure, and that migration, admixture,
or spatial structure could increase signed LD above zero.

However, under standard population genetic assumptions of constant mutation and
recombination rates and independent mutation events, migration and admixture
does not result in $\E[D]$ to be nonzero. For a given set of mutations with
differences in allele frequencies between source populations, $D$ will be
nonzero in an admixed population between those mutations even when $D$ is zero
in the sources [@Cavalli-Sforza1971-jv]. However, genome averages of allele
frequency differences in the source populations are zero, as they can be either
positive or negative, and resulting $\E[D]$ is still zero genome-wide.
Furthermore, @Sandler2020-vk used forward simulations with an explicit spatial
model and a multiplicative fitness function to explore this effect, and the
positive LD they observed under these scenarios extends over larger distances
than observed in the data. Instead, positive LD between synonymous mutations
decays to zero within just a few hundred base (Figures \@ref(fig:lddecay) and
\@ref(fig:lddecaydomain)).

### 2021-03-08 SupportingInformation.Rmd
edits

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "sipreamble.tex"
    extra_dependencies: ["float"]
bibliography: "references.bib"
title: Supporting Information for 
author:
  - Aaron P. Ragsdale^[aaronpeaceragsdale@gmail.com]
date:
  - \today
---

>>>>

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "sipreamble.tex"
    extra_dependencies: ["float"]
bibliography: "references.bib"
title: Supporting Information for "Can we distinguish model of selective interactions using linkage disequilibrium?"
author:
  - Aaron P. Ragsdale
date:
  - \today
---
